For demonstration purposes, I create a sample benchmark with 300 non-empty genes from the `pbmc4k` dataset:

```{r, include = FALSE}
pbmcs <- TENxPBMCData::TENxPBMCData("pbmc4k")
pbmcs
```

```{r}
non_empty_rows <- which(DelayedMatrixStats::rowSums2(SummarizedExperiment::assay(pbmcs)) > 0)
pbmcs_subset <- as.matrix(SummarizedExperiment::assay(pbmcs)[sample(non_empty_rows, 300), ])
model_matrix <- matrix(1, nrow = ncol(pbmcs_subset))
dim(pbmcs_subset)
```


I compare my method (in-memory and on-disk) with `DESeq2` and `edgeR`. Both are classical methods for analyzing RNA-Seq datasets and have been around for almost 10 years. Note that both tools can do a lot more than just fitting the Gamma-Poisson model, so this benchmark only serves to give a general impression of the performance.

```{r, warning=FALSE}
bench::mark(
  glmGamPoi_in_memory = {
    glmGamPoi::glm_gp(pbmcs_subset, design = model_matrix, on_disk = FALSE)
  }, glmGamPoi_on_disk = {
    glmGamPoi::glm_gp(pbmcs_subset, design = model_matrix, on_disk = TRUE)
  }, DESeq2 = suppressMessages({
    dds <- DESeq2::DESeqDataSetFromMatrix(pbmcs_subset,
                        colData = data.frame(name = seq_len(4340)),
                        design = ~ 1)
    dds <- DESeq2::estimateSizeFactors(dds, "poscounts")
    dds <- DESeq2::estimateDispersions(dds, quiet = TRUE)
    dds <- DESeq2::nbinomWaldTest(dds, minmu = 1e-6)
  }), edgeR = {
    edgeR_data <- edgeR::DGEList(pbmcs_subset)
    edgeR_data <- edgeR::calcNormFactors(edgeR_data)
    edgeR_data <- edgeR::estimateDisp(edgeR_data, model_matrix)
    edgeR_fit <- edgeR::glmFit(edgeR_data, design = model_matrix)
  }, check = FALSE
)
```

Fitting the full `pbmc4k` dataset on a modern MacBook Pro in-memory takes ~1 minute and on-disk a little over 4 minutes. Fitting the `pbmc68k` (17x the size) takes ~73 minutes (17x the time) on-disk. Fitting that dataset in-memory is not possible because it is just too big: the maximum in-memory matrix size is `2^31-1 ≈ 2.1e9` is elements, the `pbmc68k` dataset however has nearly 100 million elements more.



Comparing the results:

```{r message=FALSE, warning=FALSE}
# Results with my method
fit <- glmGamPoi::glm_gp(pbmcs_subset, design = model_matrix, on_disk = FALSE)

# DESeq2
dds <- DESeq2::DESeqDataSetFromMatrix(pbmcs_subset, 
                        colData = data.frame(name = seq_len(4340)),
                        design = ~ 1)
dds <- DESeq2::estimateSizeFactors(dds, "poscounts")
dds <- DESeq2::estimateDispersions(dds, quiet = TRUE)
dds <- DESeq2::nbinomWaldTest(dds, minmu = 1e-6)

#edgeR
edgeR_data <- edgeR::DGEList(pbmcs_subset)
edgeR_data <- edgeR::calcNormFactors(edgeR_data)
edgeR_data <- edgeR::estimateDisp(edgeR_data, model_matrix)
edgeR_fit <- edgeR::glmFit(edgeR_data, design = model_matrix)
```


```{r coefficientComparison, fig.height=5, fig.width=10, warning=FALSE, echo = FALSE}
par(mfrow = c(2, 4), cex.main = 2, cex.lab = 1.5)
plot(fit$Beta[,1], coef(dds)[,1] / log2(exp(1)), pch = 16, 
     main = "Beta Coefficients", xlab = "glmGamPoi", ylab = "DESeq2")
abline(0,1)
plot(fit$Beta[,1], edgeR_fit$unshrunk.coefficients[,1], pch = 16,
     main = "Beta Coefficients", xlab = "glmGamPoi", ylab = "edgeR")
abline(0,1)

plot(fit$Mu[,1], rowData(dds)$baseMean, pch = 16, log="xy",
     main = "Gene Mean", xlab = "glmGamPoi", ylab = "DESeq2")
abline(0,1)
plot(fit$Mu[,1], edgeR_fit$fitted.values[,1], pch = 16, log="xy",
     main = "Gene Mean", xlab = "glmGamPoi", ylab = "edgeR")
abline(0,1)

plot(fit$overdispersions, rowData(dds)$dispGeneEst, pch = 16, log="xy",
     main = "Overdispersion", xlab = "glmGamPoi", ylab = "DESeq2")
abline(0,1)
plot(fit$overdispersions, edgeR_fit$dispersion, pch = 16, log="xy",
     main = "Overdispersion", xlab = "glmGamPoi", ylab = "edgeR")
abline(0,1)

```

The inferred Beta coefficients and gene means agree well between the methods, however the overdispersion differs quite a bit. `DESeq2` has problems estimating some of the overdispersions and sets them to `1e-8`. `edgeR` only approximates the overdispersions which explains the variation around the overdispersions calculated with `glmGamPoi`.

